---
title: 'Data Manipulation'
author: "Bert Gollnick"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

# Introduction

_dplyr_ is one of the most popular and most useful packages in the R-universe. It covers all aspects of data manipulation: filtering, selection, sorting and many more.

* Objectives: data manipulation with _dplyr_
* Requirements: none

In our example we will find out most popular girl-names from 2014 and get more details of their popularity from 1880 to 2014.

# Packages

We will analyse baby names from package _babynames_. Please make sure you have installed it before trying to load it. The same holds for _dplyr_.

* _babynames_: includes data for this tutorial
* _dplyr_: includes all data manipulation 

```{r}
library(babynames)  # data for this tutorial
library(dplyr)  # data manipulation
library(ggplot2)  # visualisation of results
```

# Data Preparation

You can load the dataframe using _data_ and get an overview of structure with _tbl_df_.
```{r}
# add code here
data("babynames")
glimpse(babynames)
str(babynames)
```

There are five columns: _year_, _sex_, _name_, a count _n_, and corresponding proportion _prop_.


## Filtering

We want to keep only data from 2014 and can get this with _filter_. First, take a look at the pipeless version. The first argument is the data, followed by the filter to be applied.

```{r}
# add code here
## Getting the data from year 2014
most_popular_females = babynames%>%
  filter(year==2014)
```

We have a dataframe with all names, but only from 2014. For the remainder we will always use piping where applicable.

We also have to filter for sex, so add an additional filter.

```{r}
# add code here

```

Hint: some identical function names are part of other packages as well, e.g. _filter_ and might be masked by them. This means if you just call _filter_ the filter function from a different package is used instead of _dplyr_. Of course this does not work. Some workarounds: make sure that only _dplyr_ is loaded; make sure _dplyr_ is loaded before the other packages with identical function names. Another option to avoid trouble is to specifically define _filter_ function from _dplyr_ package by stating _dplyr::filter()_.

## Sorting

Data can be sorted with _arrange_. This function can have several sorting columns. Default-order is ascending. If descending is required, it can be done with _desc()_.

In our example data is sorted according to _sex_ and descending proportion _prop_.

```{r}
# add code here

```

## Column Selection

Often, not all columns are needed for further analysis. _select_ can be applied to reduce number of columns. It can be done by defining columns to be kept or by defining columns to be excluded.

In our example only _name_ is kept. An alternative would have been _select(-year, -sex, -n, -prop)_.

```{r}
# add code here

```

Only the Top5 will be analysed. So at the end _head_ is applied.
```{r}
# add code here

```

These are in decreasing order the most famous girlnames of 2014.

## Plot Results

Now, the development of these names should be shown. Original dataframe is filtered for these names. Function _mutate_  can create new variables.

```{r}
# add code here

```

Plot is created with _ggplot_. It shows Top5 names of 2014 and their development since 1880.
```{r}
ggplot(babies_to_plot, aes(year, prop, color = name)) +
	geom_line(size = 2) +
	xlab ("Year [-]") + 
	ylab ("Proportion [%]") + 
	ggtitle ("Most Popular Female Names in 2014, and Before") + 
	theme_bw()
```

Emma was very famous in 1880 and following decades and only recently experienced a revival. The other Top5 names got famous only recently. 

# Conclusion

We learned _dplyr_ package and some of its very important functions. There is more to discover. For a quick overview take a look at "Data Manipulation with dplyr, tidyr" cheatsheet (Help --> Cheatsheets).